Rotation Period Evolution in Low-Mass Binary Stars: The Impact of Tidal Torques and Magnetic Braking

### 2019-04-09 Paper/sync.tex
Updates from Overleaf

\newcommand{\vplanet}[0]{\texttt{VPLanet}\xspace}
\newcommand{\eqtide}[0]{\texttt{EQTIDE}\xspace}
\newcommand{\stellar}[0]{\texttt{STELLAR}\xspace}
\newcommand{\kepler}[0]{\textit{Kepler}\xspace}

@@

We examine how tides, stellar evolution, and magnetic braking shape the rotation period (P$_{rot}$) evolution of low-mass stellar binaries up to orbital periods (P$_{orb}$) of 100 d across a wide range tidal dissipation parameters using two common equilibrium tidal models. We find that many binaries with P$_{orb} \lsim 20$ d tidally-lock, and most with $P_{orb} \lsim 4$ d tidally-lock into synchronous rotation on circularized orbits. At short P$_{orb}$, tidal torques produce a population of fast rotators that single-star only models of magnetic braking fail to produce.  In many cases, we show that the competition between magnetic braking and tides produces a population of subsynchronous rotators that persists for Gyrs, even in short P$_{orb}$ binaries, qualitatively reproducing the subsynchronous eclipsing binaries (EBs) discovered in the \kepler field by \citet{Lurie2017}. Both equilibrium tidal models predict that binaries can tidally-interact out to P$_{orb} \approx 80$ d, while the CPL equilibrium tidal model predicts that binaries can tidally-lock out to P$_{orb} \approx 100$ d. Tidal torques often force the P$_{rot}$ evolution of stellar binaries to depart from the long-term magnetic braking-driven spin down experienced by single stars, revealing that P$_{rot}$ is not be a valid proxy for age in all cases, i.e. gyrochronology methods can fail unless one accounts for binarity. We suggest that accurate determinations of orbital eccentricties and P$_{rot}$ can be used to discriminate between which equilibrium tidal models best describes tidal interactions in low-mass binary stars.

@@

The long-term angular momentum evolution of low-mass (M$\lsim 1$ M$_{\odot}$) stars is controlled by magnetic braking, the torque exerted on stars due to the coupling of stellar winds to the surface magnetic field \citep{Mestel1968}. Early in stellar lifetimes, stars spin-up as they contract along the pre-main sequence.  Once stars reach the main sequence, stellar radii remain mostly constant while magnetic braking removes angular momentum from the stars, gradually spinning them down over time \citep{Skumanich1972}. Although the precise details of how magnetic braking operates are not fully known, models of magnetic braking have been used to successfully model the bulk trends of P$_{rot}$ distributions in clusters \citep[e.g. Praesepe, ][]{Matt2015,Douglas2017} and field stars \citep[e.g. the \kepler field, ][]{Reiners2012,Matt2015,vanSaders2018}. Furthermore, the magnetic braking-driven long-term spin-down of stars has been used to estimate stellar ages, a method known as gyrochronlogy \citep{Skumanich1972,Barnes2003,Barnes2007,Mamajek2008,Barnes2010}, with older stars assumed to have lost more angular momentum due to magnetic braking and therefore rotate more slowly.

@@

Here, we present a model for the angular momentum evolution of low-mass stellar binaries over their full premain and main sequence lifetimes using a realistic treatment of stellar evolution, magnetic braking, and tidal torques. We investigate under what conditions tidal-locking occurs, and how tidal torques influence rotation in stellar binaries as a function of binary P$_{orb}$ and tidal dissipation parameters for two widely-used equilibrium tidal models and two magnetic braking models.  We show how tidal torques can impact stellar rotation in binaries out to P$_{orb} = 100$ d, causing stellar rotation periods to not strongly correlate with age, making the predictions of gyrochronlogy models to potentially fail in such systems.  We describe our model in $\S$~\ref{sec:methods} and our simulation procedure in $\S$~\ref{sec:simulations}.  We discuss our results in $\S$~\ref{sec:results}, apply our model to the \kepler field in $\S$~\ref{sec:kepler}, and discuss our results' implications in $\S$~\ref{sec:discussion}.

@@

The first magnetic braking model we use was derived by \citet{Matt2015} and has been shown to successfully model the spin-down of low-mass stars across many ages in both the Praesepe cluster and in the \kepler field. This model is strongly dependent on the stellar Rossby number, $Ro = $ P$_{rot}/\tau_{cz}$, the ratio of the stellar P$_{rot}$ to the stellar convective turnover timescale, $\tau_{cz}$. The \citet{Matt2015} model predicts that below a certain $Ro$ for rapidly rotating stars, stellar magnetic activity saturates at a constant value, producing a magnetic braking torque that is directly proportional to the stellar rotation rate.  The angular momentum loss for rapidly-rotating saturated stars is given by
\begin{equation} \label{eqn:mattSat}
\frac{dJ}{dt} = -\frac{dJ}{dt}\Bigg|_0 \chi^2 \left( \frac{\omega}{\omega_{\odot}} \right) 
\end{equation}
while for more slowly-rotating unsaturated stars,
\begin{equation} \label{eqn:mattUnSat}
\frac{dJ}{dt} = -\frac{dJ}{dt}\Bigg|_0 \left( \frac{\tau_{cz}}{\tau_{cz \odot}} \right) \left( \frac{\omega}{\omega_{\odot}} \right)^3 
\end{equation}
where
\begin{equation} \label{eqn:matt0}
\frac{dJ}{dt}\Bigg|_0 = 6.3 \times 10^{30} \ \mathrm{erg} \ \left( \frac{R}{R_{\odot}} \right)^{3.1} \left( \frac{M}{M_{\odot}} \right)^{0.5}.
\end{equation}
Saturated magnetic braking occurs for $Ro \leq Ro_{\odot}/\chi$ for $\chi = 10$ where \citet{Matt2015} defines $\chi = Ro_{\odot}/Ro_{sat}$.  We adopt all model parameters given in Table 1 from \citet{Matt2015}, with the correction from \citet{Matt2019}, and compute $\tau_{cz}$ using Eqn. (36) from \citet{Cranmer2011}.

The second model we consider was developed by \citet{Reiners2012}, who modeled stellar spin-down as a function stellar magnetic field strength.  Similar to the \citet{Matt2015} model, \citet{Reiners2012} posit that magnetic braking can be divided into a unsaturated ($\omega < \omega_{crit}$) and unsaturated ($\omega \geq \omega_{crit}$) regime as follows:
\begin{equation} \label{eqn:reiners}
\begin{split}
\frac{dJ_{\star}}{dt} & = -C \left[ \omega \left(\frac{R^{16}}{M^2} \right)^{1/3} \right] \text{for $\omega \geq \omega_{crit}$} \\
\frac{dJ_{\star}}{dt} & = -C \left[ \left( \frac{\omega}{\omega_{crit}} \right)^4 \omega \left(\frac{R^{16}}{M^2} \right)^{1/3} \right] \text{for $\omega < \omega_{crit}$}.
\end{split}
\end{equation}
\citet{Reiners2012} tuned their model to match the Sun's P$_{rot}$ and the distribution of P$_{rot}$ of Gyr-old low-mass field stars and find $C = 2.66 \times 10^3 \text{ (gm$^5$ cm$^{-10}$ s$^3$)$^{1/3}$}$, $\omega_{crit} = 8.56 \times 10^{-6}\text{ s$^{-1}$}$ for $M > 0.35$ M$_{\odot}$, and for late M dwarfs with $M \leq 0.35 M_{\odot}$, $\omega_{crit} = 1.82 \times 10^{-6} \text{ s$^{-1}$}$.

@@

In Fig.~\ref{fig:stellarExample}, we plot the evolution of $R$, $r_g$, and P$_{rot}$ for 0.2 M$_{\odot}$, 0.7 M$_{\odot}$, and 1 M$_{\odot}$ mass stars, representing an M, K , and G dwarf, respectively, computed according to our stellar evolution model, \stellar, using both magnetic braking models. We assume all stars have an initial P$_{rot} = 1$ d and have an initial age of 5 Myr. All stars' radii contract along the pre-main sequence, spinning the stars up (right panel). Once the stars reach the main sequence, their structure changes slowly, allowing magnetic braking to dominate the stellar angular momentum evolution, significantly spinning-down the stars over long timescales. Both magnetic braking models produce qualitatively similar P$_{rot}$ evolution, with the \citet{Reiners2012} predicting marginally stronger torques. The $r_g$ evolution noticeably differs between the stars as the late M dwarf's (green) $r_g$ varies little as it remains fully convective, while the K and G dwarf grow a radiative core while on the pre-main sequence, decreasing $r_g$ until both reach the main sequence.

\begin{figure*}[ht]
	\includegraphics[width=\textwidth]{../Plots/stellarExample.pdf}
   \caption{Stellar $R$ (left), $r_g$ (middle), and P$_{rot}$ (right) evolution for 0.2 M$_{\odot}$ (M, green), 0.7 M$_{\odot}$ (K, orange), and 1 M$_{\odot}$ (G, blue) mass stars computed according to \stellar, our interpolation of the \citet{Baraffe2015} stellar evolution models ($\S$~\ref{sec:methods:stellar}) combined with the \citet{Matt2015} (solid) and \citet{Reiners2012} (dashed) magnetic braking equations. Each dot denotes the approximate time when each star reaches the main sequence.}%
    \label{fig:stellarExample}%
\end{figure*}

@@

Values for stellar tidal $Q$s and $\tau$s for low-mass stars are highly uncertain due to complex viscous evolution within the stars \citep{Ogilvie2007}, and can differ for stars of the same spectral class \citep{Barker2009}. These parameters can also vary as a function of stellar mass or age \citep{Bolmont2016,vanEylen2016}, likely due to low-mass stars' evolving convective regions where the tidal dissipation predominantly occurs \citep{Zahn2008}. Typical values of $Q$ and $\tau$ for Sun-like stars are estimated to be of order $Q \approx 10^6$ and $\tau \approx 0.1$ s, respectfully \citep[e.g.][]{Meibom2005,Ogilvie2007,Jackson2008}, however a range of values exist in the literature.  Therefore, we consider a wide range of tidal parameters by sampling stellar tidal $Q$s log-uniformly over $[10^4,10^8]$ and $\tau$ log-uniformly over $[10^{-2},10]$ s.  There is no general expression to compute $Q$ as a function of $\tau$, or vice versa, except in some special cases where approximations exist, e.g. Eqn. (2) from \citet{Heller2011}.

We explore the impact of both the \citet{Matt2015} and \citet{Reiners2012} magnetic braking models, but default to using the \citet{Matt2015} unless noted otherwise. All stars have an initial age of 5 Myr unless stated otherwise as by this time, the gaseous protoplanterary circumbinary disk that can drive significant dynamical evolution in the binary \citep[e.g.][]{Fleming2017} would likely have dissipated \citep{Haisch2001}.

@@

@@

\subsubsection{Torque Balance Using the \citet{Matt2015} Magnetic Braking Model} \label{sec:mattBalance}

@@

\subsubsection{Torque Balance Using the \citet{Reiners2012} Magnetic Braking Model}

The mechanism for producing subsynchronous rotation is not limited to the \citet{Matt2015} magnetic braking model, but is a generic outcome of the coupling of magnetic braking and weak tidal torques. The CTL model can only tidally-lock the P$_{orb} = 5$ d binary as the stronger magnetic braking under the \citet{Reiners2012} model forces the longer P$_{orb}$ binaries into subsynchronous rotation. Curiously, all CTL subsynchronous rotators seem to converge to P$_{rot}/$P$_{eq} \approx 1.1$.  In this state, tidal torques nearly balance magnetic braking, producing a long-lasting population of subsynchronous rotators, similar to what was observed above.  In the right panel of Fig.~\ref{fig:eqPerReiners}, we see that the net P$_{rot}$ derivatives for these binaries oscillate about 0, but are slightly negative due to the 3 simultaneous physical effects described above. This slight negative derivative is effectively negligible, however, as the torque balance will likely hold the binaries in subsynchronous rotation for 10s of Gyrs.  We note that in other exploratory simulations with larger $\tau$, we find that the tides do in fact eventually win out over magnetic braking in the long-run, driving P$_{rot}$ back towards P$_{eq}$.

\begin{figure*}
	\includegraphics[width=\textwidth]{../Plots/eqPerTwoPanelReiners.pdf}
   \caption{Same format as Fig.~\ref{fig:eqPer}, but evolved using the \citet{Reiners2012} magnetic braking model.}%
    \label{fig:eqPerReiners}%
\end{figure*}

@@

We next examine how P$_{rot}$ evolution in stellar binaries depends on P$_{orb}$ and the strength of tidal dissipation, parameterized by $Q$ and $\tau$ for the CPL and CTL models, respectfully. In Fig.~\ref{fig:qTauLock}, we bin our simulation results after the full 7 Gyr evolution by P$_{orb}$ and $Q$ or $\tau$ and compute the median P$_{orb}/$P$_{rot}$ in each bin, marginalizing over all other parameters.

@@

We compare the P$_{rot}$ distribution for tidally-interacting stellar binaries from our CPL and CTL simulations with the P$_{rot}$ distribution of single stars subject to stellar evolution and magnetic braking to gauge the impact of tidal torques on driving P$_{rot}$ distributions away from that of single stars.  We simulate 10,000 single star systems according to the evolution described in $\S$~\ref{sec:methods:stellar} with initial conditions sampled from the same mass and P$_{rot}$ distributions used for the binary simulations described $\S$~\ref{sec:methods}. In Fig.~\ref{fig:protDist}, we display P$_{rot}$ as a function of mass and age for binaries simulated using both the CPL and CTL model and for single stars. We perform this procedure for stars losing angular momentum via magnetic braking according to both the \citet{Matt2015} and \citet{Reiners2012} magnetic braking models.

\subsubsection{\citet{Matt2015} Magnetic Braking} \label{sec:mattDist}

@@

\subsubsection{\citet{Reiners2012} Magnetic Braking} \label{sec:reinersDist}

Similar to the \citet{Matt2015} simulations discussed above, the single star simulations display a clear age-dependence of P$_{rot}$. The \citet{Reiners2012} model's magnetic braking transition at $M = 0.35$ M$_{\odot}$ produces a sharp cutoff in the P$_{rot} - M$ distribution between late M dwarfs and more massive stars.  More massive stars ($M \geq 0.35$ M$_{\odot}$) follow a smooth P$_{rot} -$ age sequence, while the late M dwarfs can remain rapidly rotating if they are young and in the saturated regime or extremely slowly rotating if they are older than about 4 Gyr. As in the \citet{Matt2015} model, the only rapid rotators (P$_{rot} \lsim 20$ d) are very young stars with ages $\lsim 2$ Gyr. The presence of extremely slowly rotating late M dwarfs, absent in the predictions of the \citet{Matt2015} model but seemingly present in nature \citep[e.g.][]{Newton2018}, could be used as an observational discriminant between the \citet{Matt2015} and \citet{Reiners2012} model, but a full investigation of which magnetic braking model accurately describes angular momentum loss in low mass stars is beyond the scope of this work.

\begin{figure*}[t]
	\includegraphics[width=\textwidth]{../Plots/protDistReiners.pdf}
   \caption{Same format as Fig.~\ref{fig:protDist}, but evolved using the \citet{Reiners2012} magnetic braking model.}%
    \label{fig:protDistReiners}%
\end{figure*}

The general trends found in the coupled stellar-tidal simulations using the \citet{Matt2015} model persist in the simulations evolved using the \citet{Reiners2012} model, demonstrating that these findings are not magnetic braking model-dependant, but rather a generic result of the interaction between magnetic braking and tidal torques in low-mass binaries. Both the CPL and CTL simulations display a clear single-star magnetic braking-dominated sequence, but as before, tidally-locked binaries appear across a wide range of P$_{rot}$ and primary star masses. Both the CPL and CTL model tidally-lock many short P$_{orb}$ stars into rapid rotation across all masses, with both models predicting $23\%$ of stars have P$_{rot} < 20$ d, compared to $11\%$ of single-stars, mostly for late M dwarfs in the latter case. As shown above, the CTL model predicts that binaries can strongly tidally-interact out to long P$_{orb}$, seen out to P$_{rot} = 80$ d, while the CPL model can tidally-lock stars out to P$_{orb} \approx 100$ d. These strong tidal interactions seen both here and in $\S$~\ref{sec:mattDist} decouple P$_{rot}$ from age in binaries, especially for rapid rotators.

\subsubsection{Impact of Binarity on Gyrochronology}

Our models predict that age does not strongly correlate with P$_{rot}$ in low-mass binaries, especially for P$_{orb} < 20 $ d. Under the \citet{Matt2015} model discussed in $\S$~\ref{sec:mattDist}, the median ages and $68\%$ interval for fast rotators (P$_{rot} \lsim 20$ d) are $2.8^{+2.9}_{-1.5}$ Gyr and $3.1^{+2.6}_{-1.8}$ Gyr according to the CPL and CTL models, respectfully, compared to much younger single stars with ages of $1.3^{+0.5}_{-0.3}$ Gyr. Similarly for the models evolved with the \citet{Reiners2012} model discussed in $\S$~\ref{sec:reinersDist}, the median ages and $68\%$ interval for fast rotators are $2.6^{+2.8}_{-1.3}$ Gyr and $2.7^{+2.7}_{-1.4}$ Gyr according to the CPL and CTL models, respectfully, compared to single stars with ages of $1.7^{+0.9}_{-0.5}$ Gyr, We highlight this stark dichotomy between the age distribution of single and double stars in Fig.~\ref{fig:protAgeHist} by plotting a histogram of system ages from Fig.~\ref{fig:protDist} for single or primary stars in binaries with P$_{rot} < 20$ d.

@@

% In binaries, however, tidal torques can lock stellar P$_{rot}$ to P$_{eq}$, especially at short P$_{orb}$ where tidal torques are stronger, creating a long-lasting population of fast rotators with a wide age range. 
% This feature is not due to long-term spin-down via magnetic braking as this population lies above the old edge of single star sequence. Instead

>>>>

\newcommand{\xxx}[1]{{\textbf{#1}}}
\newcommand{\vplanet}[0]{\texttt{VPLanet}\xspace}
\newcommand{\eqtide}[0]{\texttt{EQTIDE}\xspace}
\newcommand{\stellar}[0]{\texttt{STELLAR}\xspace}
\newcommand{\kepler}[0]{\textit{Kepler}\xspace}

@@

We examine how tides, stellar evolution, and magnetic braking shape the rotation period (P$_{rot}$) evolution of low-mass stellar binaries up to orbital periods (P$_{orb}$) of 100 d across a wide range tidal dissipation parameters using two common equilibrium tidal models. We find that many binaries with P$_{orb} \lsim 20$ d tidally-lock, and most with $P_{orb} \lsim 4$ d tidally-lock into synchronous rotation on circularized orbits. At short P$_{orb}$, tidal torques produce a population of fast rotators that single-star only models of magnetic braking fail to produce.  In many cases, we show that the competition between magnetic braking and tides produces a population of subsynchronous rotators that persists for Gyrs, even in short P$_{orb}$ binaries, qualitatively reproducing the subsynchronous eclipsing binaries (EBs) discovered in the \kepler field by \citet{Lurie2017}. Both equilibrium tidal models predict that binaries can tidally-interact out to P$_{orb} \approx 80$ d, while the \xxx{Constant Phase Lag tidal} model predicts that binaries can tidally-lock out to P$_{orb} \approx 100$ d. Tidal torques often force the P$_{rot}$ evolution of stellar binaries to depart from the long-term magnetic braking-driven spin down experienced by single stars, revealing that P$_{rot}$ is not be a valid proxy for age in all cases, i.e. gyrochronology methods can fail unless one accounts for binarity. We suggest that accurate determinations of orbital eccentricties and P$_{rot}$ can be used to discriminate between which equilibrium tidal models best describes tidal interactions in low-mass binary stars.

@@

The long-term angular momentum evolution of low-mass (M$\lsim 1$ M$_{\odot}$) stars is controlled by magnetic braking, the torque exerted on stars due to the coupling of stellar winds to the surface magnetic field \citep{Mestel1968}. Early in stellar lifetimes, stars spin-up as they contract along the pre-main sequence.  Once stars reach the main sequence, stellar radii remain mostly constant while magnetic braking removes angular momentum from the stars, gradually spinning them down over time \citep{Skumanich1972}. Although the precise details of how magnetic braking operates are not fully known, models of magnetic braking have been used to successfully model the bulk trends of P$_{rot}$ distributions in clusters \textbf{\citep[e.g. Praesepe, ][]{Reiners2012,Matt2015,Douglas2017}} and field stars \citep[e.g. the \kepler field, ][]{Matt2015,vanSaders2018}. Furthermore, the magnetic braking-driven long-term spin-down of stars has been used to estimate stellar ages, a method known as \xxx{gyrochronology} \citep{Skumanich1972,Barnes2003,Barnes2007,Mamajek2008,Barnes2010}, with older stars assumed to have lost more angular momentum due to magnetic braking and therefore rotate more slowly.

@@

Here, we present a model for the angular momentum evolution of low-mass stellar binaries over their full premain and main sequence lifetimes using a realistic treatment of stellar evolution, magnetic braking, and tidal torques. We investigate under what conditions tidal-locking occurs, and how tidal torques influence rotation in stellar binaries as a function of binary P$_{orb}$ and tidal dissipation parameters for two widely-used equilibrium tidal models and two magnetic braking models.  We show how tidal torques can impact stellar rotation in binaries out to P$_{orb} = 100$ d, causing stellar rotation periods to not strongly correlate with age, making the predictions of \xxx{gyrochronology} models to potentially fail in such systems.  We describe our model in $\S$~\ref{sec:methods} and our simulation procedure in $\S$~\ref{sec:simulations}.  We discuss our results in $\S$~\ref{sec:results}, apply our model to the \kepler field in $\S$~\ref{sec:kepler}, and discuss our results' implications in $\S$~\ref{sec:discussion}.

@@

\xxx{We simulate magnetic braking using the model derived by \citet{Matt2015} as this formalism} has been shown to successfully model the spin-down of low-mass stars across many ages in both the Praesepe cluster and in the \kepler field. This model is strongly dependent on the stellar Rossby number, $Ro = $ P$_{rot}/\tau_{cz}$, the ratio of the stellar P$_{rot}$ to the stellar convective turnover timescale, $\tau_{cz}$. The \citet{Matt2015} model predicts that below a certain $Ro$ for rapidly rotating stars, stellar magnetic activity saturates at a constant value, producing a magnetic braking torque that is directly proportional to the stellar rotation rate.  The angular momentum loss for rapidly-rotating saturated stars is given by
\begin{equation} \label{eqn:mattSat}
\frac{dJ}{dt} = -\frac{dJ}{dt}\Bigg|_0 \chi^2 \left( \frac{\omega}{\omega_{\odot}} \right) 
\end{equation}
\xxx{while for more slowly-rotating unsaturated stars,}
\begin{equation} \label{eqn:mattUnSat}
\frac{dJ}{dt} = -\frac{dJ}{dt}\Bigg|_0 \left( \frac{\tau_{cz}}{\tau_{cz \odot}} \right)^2 \left( \frac{\omega}{\omega_{\odot}} \right)^3
\end{equation}
where
\begin{equation} \label{eqn:matt0}
\frac{dJ}{dt}\Bigg|_0 = 6.3 \times 10^{30} \ \mathrm{erg} \ \left( \frac{R}{R_{\odot}} \right)^{3.1} \left( \frac{M}{M_{\odot}} \right)^{0.5}.
\end{equation}
Saturated magnetic braking occurs for $Ro \leq Ro_{\odot}/\chi$ for $\chi = 10$ where \citet{Matt2015} defines $\chi = Ro_{\odot}/Ro_{sat}$.  We adopt all model parameters given in Table 1 from \citet{Matt2015}, with the correction from \citet{Matt2019}, and compute $\tau_{cz}$ using Eqn. (36) from \citet{Cranmer2011}.

%The second model we consider was developed by \citet{Reiners2012}, who modeled stellar spin-down as a function stellar magnetic field strength.  Similar to the \citet{Matt2015} model, \citet{Reiners2012} posit that magnetic braking can be divided into a unsaturated ($\omega < \omega_{crit}$) and unsaturated ($\omega \geq \omega_{crit}$) regime as follows:
%\begin{equation} \label{eqn:reiners}
%\begin{split}
%\frac{dJ_{\star}}{dt} & = -C \left[ \omega \left(\frac{R^{16}}{M^2} \right)^{1/3} \right] \text{for $\omega \geq \omega_{crit}$} \\
%\frac{dJ_{\star}}{dt} & = -C \left[ \left( \frac{\omega}{\omega_{crit}} \right)^4 \omega \left(\frac{R^{16}}{M^2} \right)^{1/3} \right] %\text{for $\omega < \omega_{crit}$}.
%\end{split}
%\end{equation}
%\citet{Reiners2012} tuned their model to match the Sun's P$_{rot}$ and the distribution of P$_{rot}$ of Gyr-old low-mass field stars and find $C = 2.66 \times 10^3 \text{ (gm$^5$ cm$^{-10}$ s$^3$)$^{1/3}$}$, $\omega_{crit} = 8.56 \times 10^{-6}\text{ s$^{-1}$}$ for $M > 0.35$ M$_{\odot}$, and for late M dwarfs with $M \leq 0.35 M_{\odot}$, $\omega_{crit} = 1.82 \times 10^{-6} \text{ s$^{-1}$}$.

@@

In Fig.~\ref{fig:stellarExample}, we plot the evolution of $R$, $r_g$, and P$_{rot}$ for 0.2 M$_{\odot}$, 0.7 M$_{\odot}$, and 1 M$_{\odot}$ mass stars, representing an M, K , and G dwarf, respectively, computed according to our stellar evolution model, \stellar, using both magnetic braking models. We assume all stars have an initial P$_{rot} = 1$ d and have an initial age of 5 Myr. All stars' radii contract along the pre-main sequence, spinning the stars up (right panel). Once the stars reach the main sequence, their structure changes slowly, allowing magnetic braking to dominate the stellar angular momentum evolution, significantly spinning-down the stars over long timescales. The $r_g$ evolution noticeably differs between the stars as the late M dwarf's (green) $r_g$ varies little as it remains fully convective, while the K and G dwarf grow a radiative core while on the pre-main sequence, decreasing $r_g$ until both reach the main sequence.

% Cut:
% Both magnetic braking models produce qualitatively similar P$_{rot}$ evolution, with the \citet{Reiners2012} predicting marginally stronger torques.

\begin{figure*}[ht]
	\includegraphics[width=\textwidth]{../Plots/stellarExample.pdf}
   \caption{Stellar $R$ (left), $r_g$ (middle), and P$_{rot}$ (right) evolution for 0.2 M$_{\odot}$ (M, green), 0.7 M$_{\odot}$ (K, orange), and 1 M$_{\odot}$ (G, blue) mass stars computed according to \stellar, our interpolation of the \citet{Baraffe2015} stellar evolution models ($\S$~\ref{sec:methods:stellar}) combined with the \xxx{\citet{Matt2015} magnetic braking model}. Each dot denotes the approximate time when each star reaches the main sequence.}%
    \label{fig:stellarExample}%
\end{figure*}

@@

Values for stellar tidal $Q$s and $\tau$s for low-mass stars are highly uncertain due to complex viscous evolution within the stars \citep{Ogilvie2007}, and can differ for stars of the same spectral class \citep{Barker2009}. These parameters can also vary as a function of stellar mass or age \citep{Bolmont2016,vanEylen2016}, likely due to low-mass stars' evolving convective regions where the tidal dissipation predominantly occurs \citep{Zahn2008}. Typical values of $Q$ and $\tau$ for Sun-like stars are estimated to be of order $Q \approx 10^6$ and $\tau \approx 0.1$ s, \xxx{respectively} \citep[e.g.][]{Meibom2005,Ogilvie2007,Jackson2008}, however a range of values exist in the literature.  Therefore, we consider a wide range of tidal parameters by sampling stellar tidal $Q$s log-uniformly over $[10^4,10^8]$ and $\tau$ log-uniformly over $[10^{-2},10]$ s.  There is no general expression to compute $Q$ as a function of $\tau$, or vice versa, except in some special cases where approximations exist, e.g. Eqn. (2) from \citet{Heller2011}. \xxx{All stars have an initial age of 5 Myr unless stated otherwise as by this time, the gaseous protoplanterary circumbinary disk that can drive significant dynamical evolution in the binary \citep[e.g.][]{Fleming2017} would likely have dissipated \citep{Haisch2001}.}

@@

% cut:
% We explore the impact of both the \citet{Matt2015} and \citet{Reiners2012} magnetic braking models, but default to using the \citet{Matt2015} unless noted otherwise.

@@

\subsubsection{\xxx{Torque Balance}} \label{sec:mattBalance}

@@

% Cut as per ref's suggestion
%\subsubsection{Torque Balance Using the \citet{Reiners2012} Magnetic Braking Model}

%The mechanism for producing subsynchronous rotation is not limited to the \citet{Matt2015} magnetic braking model, but is a generic outcome of the coupling of magnetic braking and weak tidal torques. The CTL model can only tidally-lock the P$_{orb} = 5$ d binary as the stronger magnetic braking under the \citet{Reiners2012} model forces the longer P$_{orb}$ binaries into subsynchronous rotation. Curiously, all CTL subsynchronous rotators seem to converge to P$_{rot}/$P$_{eq} \approx 1.1$.  In this state, tidal torques nearly balance magnetic braking, producing a long-lasting population of subsynchronous rotators, similar to what was observed above.  In the right panel of Fig.~\ref{fig:eqPerReiners}, we see that the net P$_{rot}$ derivatives for these binaries oscillate about 0, but are slightly negative due to the 3 simultaneous physical effects described above. This slight negative derivative is effectively negligible, however, as the torque balance will likely hold the binaries in subsynchronous rotation for 10s of Gyrs.  We note that in other exploratory simulations with larger $\tau$, we find that the tides do in fact eventually win out over magnetic braking in the long-run, driving P$_{rot}$ back towards P$_{eq}$.

%\begin{figure*}
%	\includegraphics[width=\textwidth]{../Plots/eqPerTwoPanelReiners.pdf}
%   \caption{Same format as Fig.~\ref{fig:eqPer}, but evolved using the \citet{Reiners2012} magnetic braking model.}%
%    \label{fig:eqPerReiners}%
%\end{figure*}

@@

We next examine how P$_{rot}$ evolution in stellar binaries depends on P$_{orb}$ and the strength of tidal dissipation, parameterized by $Q$ and $\tau$ for the CPL and CTL models, \xxx{respectively}. In Fig.~\ref{fig:qTauLock}, we bin our simulation results after the full 7 Gyr evolution by P$_{orb}$ and $Q$ or $\tau$ and compute the median P$_{orb}/$P$_{rot}$ in each bin, marginalizing over all other parameters.

@@

We compare the P$_{rot}$ distribution for tidally-interacting stellar binaries from our CPL and CTL simulations with the P$_{rot}$ distribution of single stars subject to stellar evolution and magnetic braking to gauge the impact of tidal torques on driving P$_{rot}$ distributions away from that of single stars.  We simulate 10,000 single star systems according to the evolution described in $\S$~\ref{sec:methods:stellar} with initial conditions sampled from the same mass and P$_{rot}$ distributions used for the binary simulations described $\S$~\ref{sec:methods}. In Fig.~\ref{fig:protDist}, we display P$_{rot}$ as a function of mass and age for binaries simulated using both the CPL and CTL model and for single stars.

% cut
%  We perform this procedure for stars losing angular momentum via magnetic braking according to both the \citet{Matt2015} and \citet{Reiners2012} magnetic braking models.

%\subsubsection{\citet{Matt2015} Magnetic Braking} \label{sec:mattDist}

@@

% cut as per ref's suggestion
%\subsubsection{\citet{Reiners2012} Magnetic Braking} \label{sec:reinersDist}

%Similar to the \citet{Matt2015} simulations discussed above, the single star simulations display a clear age-dependence of P$_{rot}$. The \citet{Reiners2012} model's magnetic braking transition at $M = 0.35$ M$_{\odot}$ produces a sharp cutoff in the P$_{rot} - M$ distribution between late M dwarfs and more massive stars.  More massive stars ($M \geq 0.35$ M$_{\odot}$) follow a smooth P$_{rot} -$ age sequence, while the late M dwarfs can remain rapidly rotating if they are young and in the saturated regime or extremely slowly rotating if they are older than about 4 Gyr. As in the \citet{Matt2015} model, the only rapid rotators (P$_{rot} \lsim 20$ d) are very young stars with ages $\lsim 2$ Gyr. The presence of extremely slowly rotating late M dwarfs, absent in the predictions of the \citet{Matt2015} model but seemingly present in nature \citep[e.g.][]{Newton2018}, could be used as an observational discriminant between the \citet{Matt2015} and \citet{Reiners2012} model, but a full investigation of which magnetic braking model accurately describes angular momentum loss in low mass stars is beyond the scope of this work.

%\begin{figure*}[t]
%	\includegraphics[width=\textwidth]{../Plots/protDistReiners.pdf}
%   \caption{Same format as Fig.~\ref{fig:protDist}, but evolved using the \citet{Reiners2012} magnetic braking model.}%
%    \label{fig:protDistReiners}%
%\end{figure*}

%The general trends found in the coupled stellar-tidal simulations using the \citet{Matt2015} model persist in the simulations evolved using the \citet{Reiners2012} model, demonstrating that these findings are not magnetic braking model-dependant, but rather a generic result of the interaction between magnetic braking and tidal torques in low-mass binaries. Both the CPL and CTL simulations display a clear single-star magnetic braking-dominated sequence, but as before, tidally-locked binaries appear across a wide range of P$_{rot}$ and primary star masses. Both the CPL and CTL model tidally-lock many short P$_{orb}$ stars into rapid rotation across all masses, with both models predicting $23\%$ of stars have P$_{rot} < 20$ d, compared to $11\%$ of single-stars, mostly for late M dwarfs in the latter case. As shown above, the CTL model predicts that binaries can strongly tidally-interact out to long P$_{orb}$, seen out to P$_{rot} = 80$ d, while the CPL model can tidally-lock stars out to P$_{orb} \approx 100$ d. These strong tidal interactions seen both here and in $\S$~\ref{sec:mattDist} decouple P$_{rot}$ from age in binaries, especially for rapid rotators.

%\subsubsection{Impact of Binarity on Gyrochronology}

Our models predict that age does not strongly correlate with P$_{rot}$ in low-mass binaries, especially for P$_{orb} < 20 $ d. \xxx{The median ages and $68\%$ interval for fast rotators (P$_{rot} \lsim 20$ d) are $2.8^{+2.9}_{-1.5}$ Gyr and $3.1^{+2.6}_{-1.8}$ Gyr according to the CPL and CTL models, respectively}, compared to much younger single stars with ages of $1.3^{+0.5}_{-0.3}$ Gyr. We highlight this stark dichotomy between the age distribution of single and double stars in Fig.~\ref{fig:protAgeHist} by plotting a histogram of system ages from Fig.~\ref{fig:protDist} for single or primary stars in binaries with P$_{rot} < 20$ d.

% cut 
% Similarly for the models evolved with the \citet{Reiners2012} model discussed in $\S$~\ref{sec:reinersDist}, the median ages and $68\%$ interval for fast rotators are $2.6^{+2.8}_{-1.3}$ Gyr and $2.7^{+2.7}_{-1.4}$ Gyr according to the CPL and CTL models, respectfully, compared to single stars with ages of $1.7^{+0.9}_{-0.5}$ Gyr,

@@

% cut
% In binaries, however, tidal torques can lock stellar P$_{rot}$ to P$_{eq}$, especially at short P$_{orb}$ where tidal torques are stronger, creating a long-lasting population of fast rotators with a wide age range. 
% This feature is not due to long-term spin-down via magnetic braking as this population lies above the old edge of single star sequence. Instead 
% and Fig.~\ref{fig:protDistReiners}.
